---
title: War threats and language
subtitle: "PSCI 2227: War and State Development"
format:
  clean-revealjs:
    echo: true
    fig-align: center
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
author:
  - name: "Prof. Brenton Kenkel"
    email: brenton.kenkel@gmail.com
    affiliations: Vanderbilt University
date: 2026-04-01
---

## Setup {visibility="hidden"}

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false
#| cache: false

library(tidyverse)
library(countrycode)

dicl <- read_tsv("dicl.csv", show_col_types = FALSE)

ling_common <- dicl |>
  filter(iso3_i == iso3_j) |>
  transmute(iso3 = iso3_i, country = country_i, cnl, csl, cal)

map_regions <- tibble(region = unique(map_data("world")$region)) |>
  mutate(iso3 = countrycode(region, "country.name", "iso3c", warn = FALSE))

country_data <- map_regions |>
  left_join(ling_common, by = "iso3")

csl_fill <- scale_fill_manual(
  name = "CSL",
  values = c(
    "< 0.5" = "#FFFFB2",
    "0.5–0.75" = "#ABD0E6",
    "0.75–1.0" = "#4A93C1",
    "> 1.0" = "#08306B",
    "Unavailable" = "white"
  )
)

map_base <- map_data("world") |>
  filter(region != "Antarctica") |>
  left_join(country_data, by = "region") |>
  mutate(
    csl_bin = cut(
      csl,
      breaks = c(-Inf, 0.5, 0.75, 1.0, Inf),
      labels = c("< 0.5", "0.5–0.75", "0.75–1.0", "> 1.0")
    ) |> fct_na_value_to_level("Unavailable")
  )
```

## Recap

**Last time.** Nation-building through war (Sambanis, Skaperdas, Wohlforth).

- Sub-national identities undermine state capacity
- Victory in war $\rightarrow$ higher national status $\rightarrow$ more national identification
- Rulers may start wars *because* of the nation-building payoff

**Today.** Darden and Mylonas, "Threats to Territorial Integrity, National Mass Schooling, and Linguistic Commonality."

- War isn't the only way to build a nation --- mass schooling is another
- External threats drive the *decision* to invest in nation-building

## Today's agenda

1. The puzzle: why so much variation in linguistic commonality?
2. Theory: external threats and incentives to nation-build
3. Evidence: regional patterns and case studies


# Theory

## Linguistic commonality and nationalism

Impossible to measure "nationalism" directly

Shared language is a reasonable (though imperfect) proxy for social scientists

Basic logic of focusing on language:

- *Imagined community* --- people who've never met feel a shared fate
- A common language is close to a prerequisite for this
  - Shared media, shared literature, shared political discourse
  - Hard to imagine community without communication
- Anderson: spread of nationalism largely coincident w/ print media

## Regional variation in linguistic commonality

```{r}
#| label: map-csl
#| echo: false
#| fig-width: 11
#| fig-height: 6

map_base |>
  ggplot(aes(long, lat, group = group)) +
  geom_polygon(aes(fill = csl_bin), color = "gray30", linewidth = 0.1) +
  csl_fill +
  coord_fixed(1.3, ylim = c(-60, 85)) +
  theme_void() +
  labs(caption = "Source: DICL v2 (Gurevich et al., 2024) — Common Spoken Language index (L1 + L2)") +
  theme(legend.position = "bottom")
```

## Non-strategic explanations

**Initial diversity.**
Does linguistic diversity now just reflect the linguistic diversity of the past?

**Duration of rule.**
Do you just need the same regime in place for a long time to produce linguistic commonality?

**State capacity.**
Do you need the state to have high extractive capacity in the first place to have a national language?

## Problems with the non-strategic explanations

**Initial diversity.**

- Around 1300, virtually *all* regions had high linguistic diversity
- Even European states were linguistically diverse until recently

**Duration of rule.**

- Long rule, low penetration of colonial language: Macau, Philippines
- Short rule, high penetration of new language: Hebrew in Israel

**State capacity.**

- High capacity, low commonality: Canada, China
- Low capacity, high commonality: Tanzania, Indonesia, Central America

## Linguistic commonality as a strategic choice

Darden and Mylonas: linguistic commonality results in part from **deliberate decisions** by states --- nation-building through national schooling

Puzzle from the SSW perspective: why wouldn't *every* state do this?

. . .

Key insights:

1. It's costly to set up a national education system
2. The benefits vary depending on threat conditions
   - High threat $\leadsto$ need to provide for national defense
   - Low threat $\leadsto$ maybe better to tax less + ask less of public

## Alternatives

::: {.callout-note title="Discussion question"}
We've seen two accounts of how a state can build nationalism:

1. Sambanis et al: Victory on the battlefield
2. Darden and Mylonas: National schooling

What other mechanisms could the state use to build national sentiment?
When would they be preferable to these?
:::

## The hypotheses

**High external threat $\rightarrow$ mass national schooling $\rightarrow$ greater linguistic commonality**

- States in competitive environments invest in nation-building
- Produces durable linguistic homogeneity

**Low external threat $\rightarrow$ no incentive to nation-build $\rightarrow$ lower linguistic commonality**

- Costs of nation-building more likely to outweigh benefits
- Education left to missionaries, churches, or other non-state actors
- Result: diverse educational systems, multiple languages


# Evidence

## Regional patterns

```{r}
#| label: regional-bar
#| echo: false

tibble(
  region = c("Middle East, North Africa", "Europe", "Balkans", "East Asia",
             "South America", "Former Soviet Union", "Central America\nand Caribbean",
             "Southeast Asia", "South Asia", "Sub-Saharan Africa"),
  pct = c(94, 93, 93, 91, 89, 86, 86, 74, 50, 45)
) |>
  mutate(region = fct_reorder(region, pct)) |>
  ggplot(aes(x = pct, y = region)) +
  geom_col(fill = "#4A93C1") +
  geom_text(aes(label = paste0(pct, "%")), hjust = -0.2, size = 4.5) +
  scale_x_continuous(limits = c(0, 105), expand = c(0, 0)) +
  labs(
    x = "Avg. % speaking dominant common language",
    y = NULL,
    caption = "Source: Darden and Mylonas (2016), Table 1"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )
```

## Regional patterns and the theory

```{r}
#| label: map-csl-2
#| echo: false
#| fig-width: 10
#| fig-height: 5.5

map_base |>
  ggplot(aes(long, lat, group = group)) +
  geom_polygon(aes(fill = csl_bin), color = "gray30", linewidth = 0.1) +
  csl_fill +
  coord_fixed(1.3, ylim = c(-60, 85)) +
  theme_void() +
  labs(caption = "Source: DICL v2 (Gurevich et al., 2024) — Common Spoken Language index (L1 + L2)")
```

High commonality where there was high interstate competition (Europe, East Asia); low where borders were externally guaranteed (Sub-Saharan Africa)

## Sub-Saharan Africa

```{r}
#| label: map-csl-ssa
#| echo: false
#| fig-width: 10
#| fig-height: 5.5

map_base |>
  ggplot(aes(long, lat, group = group)) +
  geom_polygon(aes(fill = csl_bin), color = "gray30", linewidth = 0.1) +
  csl_fill +
  coord_fixed(1.3, xlim = c(-20, 55), ylim = c(-37, 20)) +
  theme_void() +
  labs(caption = "Source: DICL v2 (Gurevich et al., 2024) — Common Spoken Language index (L1 + L2)")
```

## Why Sub-Saharan Africa is so low

The D&M explanation:

- Treaty of Berlin (1884--85): Great Powers "froze" state borders
- Post-colonial African elites colluded in 1964 to preserve existing borders
- Result: near-absence of interstate war, boundary change, and external threat


Without external threats, no strong incentive to nation-build

- Education often left to missionaries --- who taught in *local* languages, not national ones
- Multiple languages cultivated, reinforcing ethnic and regional identities
- Colonial and post-colonial decisions to delegate schooling $\rightarrow$ linguistic heterogeneity

## Case study: Indonesia (high threat)

Indonesia at independence (1949):

- 75.5 million people across 18,000 islands
- 750+ languages spoken; most common (Javanese) was a minority language
- Bahasa Indonesian --- the language of the nationalist movement --- spoken by under 5%
- GDP per capita: \$803; literacy: 15--20%

Not an obvious candidate for linguistic homogenization --- yet it happened

## Indonesia's external threats

**(a) Threat of military conquest**

- Won independence through war against the Dutch and British
- Borders remained contested; boundary disputes resolved by military action


**(b) Threat of externally supported secession**

- Dutch promoted "Pasundan" state (Balinese, Ambonese, Moluccan nationalism)
- Japanese occupation had mobilized ethnic divisions
- Cold War fears: links between ethnic Chinese and PRC; rivalry with Malaysia
- Post-colonial meddling by former imperial powers

## Indonesia's response: mass schooling in Indonesian

Indonesia pursued nation-building through schools

- 1950s: mass schooling and literacy campaign --- adult education, "people's libraries," thousands of new schools
- Indonesian established as the sole national language of instruction (beyond Grade 3)
- By 1952--53: 6.4 million of 12.3 million school-age children enrolled (52%)

Results:

- Literacy: 42.9% (1961) $\rightarrow$ 59.6% (1971) $\rightarrow$ 81.5% (1990) $\rightarrow$ 90%+ (2004)
- By 2010: 88% of 237 million people were Indonesian speakers
- Linguists describe an "epidemic language shift" --- Indonesian as a "killer language"

## Difficulty: Southeast Asia

```{r}
#| label: map-csl-sa
#| echo: false
#| fig-width: 7
#| fig-height: 5

map_base |>
  ggplot(aes(long, lat, group = group)) +
  geom_polygon(aes(fill = csl_bin), color = "gray30", linewidth = 0.1) +
  csl_fill +
  coord_fixed(1.3, xlim = c(60, 130), ylim = c(-10, 40)) +
  theme_void() +
  labs(caption = "DICL v2 — CSL (L1 + L2)")
```

What explains the variation within Southeast Asia?
Is it *really* related to external threats?

## Difficulty: Latin America

::: {.columns}
:::: {.column width="45%"}
Why is Latin America uniformly high?

- Borders have been very stable since the 1830s --- similar to Africa
- Interstate war has been rare
- Thies: variation *within* region in rivalry
- Yet linguistic commonality is very high throughout the region
::::
:::: {.column width="55%"}
```{r}
#| label: map-csl-latam
#| echo: false
#| fig-width: 7
#| fig-height: 5

map_base |>
  ggplot(aes(long, lat, group = group)) +
  geom_polygon(aes(fill = csl_bin), color = "gray30", linewidth = 0.1) +
  csl_fill +
  coord_fixed(1.3, xlim = c(-120, -30), ylim = c(-57, 35)) +
  theme_void() +
  labs(caption = "DICL v2 — CSL (L1 + L2)")
```
::::
:::

## Holding alternatives to a higher standard?

**(c) D&M seem to hold their "alternative explanations" to a higher evidentiary standard than their own theory**

Their argument structure:

- Initial diversity "doesn't explain it" because Europe was diverse too
  - But this is about *relative* levels --- some places *were* more diverse
- State capacity "doesn't explain it" because weak states can homogenize
  - But this doesn't mean capacity is *irrelevant*
- Duration of rule "doesn't explain it" because of counterexamples
  - But there are counterexamples to the threat theory too


# Wrapping up

## What we did today

1. The puzzle: wide cross-national variation in linguistic commonality
   - Can't be fully explained by initial diversity, state capacity, or duration of rule

2. D&M's theory: external threats drive nation-building
   - Threatened states invest in mass schooling with national content
   - Secure states outsource education $\rightarrow$ persistent heterogeneity
   - Mechanism: national schools

3. Evidence
   - Regional patterns align broadly with the theory
   - Indonesia (high threat) vs. Congo/Zambia (low threat) as key comparison
   - Some puzzles remain: Latin America, Southeast Asia

## To do for next time

**Reminder:** No lectures next week (Apr 6--8). Final paper consultations instead.

<https://calendly.com/brenton-kenkel/wasd-paper-consultations>

**Monday, Apr 13.** Read Brewer, *The Sinews of Power*.
